GPU Acceleration of Edge-Based Motion Detection and Machine Learning-Aided Facial Recognition with NVIDIA CUDA

### 2015-12-14 documents/paper/paper.tex
Added facial recognition section + some Python + future work in conclusion

\lstset{frame=tb,
	language=C,
	aboveskip=3mm,
	belowskip=3mm,
	showstringspaces=false,
	columns=flexible,
	basicstyle={\small\ttfamily},
	numbers=none,
	numberstyle=\tiny\color{gray},
	keywordstyle=\color{blue},
	commentstyle=\color{dkgreen},
	stringstyle=\color{mauve},
	breaklines=true,
	breakatwhitespace=true,
	tabsize=3
}

@@

\title{GPU Acceleration of Machine Learning-Aided Facial Recognition and Edge-Based Motion Detection with NVIDIA CUDA}

@@

\begin{abstract}
GPUs provide a powerful platform for parallel computations on large data inputs, namely images. In this paper, we explore a GPU-based implementation of a simplified adaptation of existing edge detection algorithms fast enough to operate on frames of a continuous video stream in real-time. We also demonstrate practical applications of edge detection, including an edge-based method for motion detection estimation. Finally, we demonstrate the speedups of over 2x we achieve with GPU parallelism, as compared to a reference serial CPU-based implementation.
\end{abstract}

\section{Introduction}
Graphics processing units (GPUs) are rapidly gaining popularity as a platform for parallelized computations on massive sets of data. Since much of the computations in image processing and computer vision are easily parallelized, graphics operations on GPUs achieve significant speedups compared to those done on their serial, CPU counterparts. Further, SDKs like the NVIDIA CUDA framework provide developers easy APIs to take advantage of the parallel computing power of GPUs. We take full advantage of the computational benefits of GPUs by implementing edge detection and motion detection algorithms in CUDA C.
\par In this paper, we first detail the theory for our edge detection and motion detection algorithms. Then, we describe the GPU code implementation of those algorithms with NVIDIA CUDA. Finally, we present speedup results achieved by our CUDA implementation with respect to a reference serial, CPU implmentation.

\subsection{GPU Computation and the NVIDIA CUDA Framework}
Formally, CUDA (Compute Unified Device Architecture) is a parallel computing platform and programming model that exposes familiar C-based APIs for parallelized computations. CUDA is NVIDIA's platform for general-purpose computing on graphics processing units (commonly, GPGPU or GP$^2$U), or the use of a GPU for computations traditionally handled by the CPU. Generally, GPGPU is used to exploit the improved multithreaded performance and raw floating-point computational ability of GPUs over CPUs. For example, on modern hardware, an NVIDIA GeForce GTX 970 (1664 CUDA cores) exhibits peak single-precision floating point performance of nearly 3500 GFLOPS (floating-point operations per second), while an Intel Core i7 4790K (4C, 8T) achieves 100 GFLOPS. Our goal is to demonstrate the performance of GPU computing by solving a handful of existing problems in computer vision with CUDA: namely, edge detection and motion detection.

\subsection{OpenCV and Python}
While CUDA is an excelent language for maximizing performance, it could be infeasible due to inexperience with C, time constraints, or lack of an NVIDIA GPU. To address these constraints, we also implemented our algorithms in Python using OpenCV. Python was chosen because it is very easy for inexperienced programmers to use and OpenCV was chosen because it offers many common image processing algorithms in a variety of languages.

\subsection{Motivation}
GPU parallelization is a powerful tool for accelerating image processing tasks. The motivation for our CUDA motion detection algorithm stems from the ease of creating a GPU implementation, due in large part to the simplicity of the CUDA APIs, and the diverse applications of motion detection. For example, motion detection is applied in real-time traffic analysis, vision-based security, and consumer technology (e.g. motion-triggered home automation solutions). Further, since our computer vision algorithm calculates the location of motion in a frame based on detected edges, our approach is both more accurate and supplemented with more specific data. Therefore, we view an opportunity use CUDA to explore the speedup possible by using a GPU to accelerate this common task in computer vision.

@@

\section{Facial Recognition Algorithm}
Placeholder text

\subsection{Placeholder}
Placeholder text

\subsection{Placeholder}
Placeholder text

\subsection{GPU Parallelization and CUDA Implementation}
Placeholder text

\section{Edge Detection Algorithm}
Edge detection is the process of determining the locations of boundaries that separate objects, or sections of objects, in an image. This edge data simplifies the features of an image to only its basic outlines, which makes it a convenient input for many other problems in computer vision, including motion detection and tracking, object classification, three-dimensional reconstruction, and others. The edge detection algorithm we present is composed of three stages:
\begin{enumerate}
	\item Reduce the amount of high-frequency noise from the image using a two-dimensional low pass filter.
	\item Determine the two-dimensional gradient of the filtered image by applying partial derivatives in both the horizontal and vertical directions.
	\item Classify edges by suppressing surrounding pixels in the gradient direction whose magnitude is lesser, and applying selective thresholding to a limited apron around each selected pixel.
\end{enumerate}
Edges correspond to the points on the image that exhibit a high gradient magnitude. The edge data from our algorithm is then used as input to our motion detection algorithm.

@@

\subsection{GPU Parallelization and CUDA Implementation}
We begin with a parallelized implementation of separable convolution, since the same convolution operation is applied for both the initial Gaussian low pass filter and the edge detecting Sobel filter. Our parallel implementation involves first independently computing the convolution with the horizontal filter ($h_2$ in Equation \eqref{separable-convolution-definition}), then convolving that result with the vertical filter $h_1$. In computing the output $y[m, n]$ at every index $(i, j)$, we launch a kernel with 16 threads per block, where the total number of blocks along one dimension is equal to the size of that dimension divided by 16. For our 5312 x 2988 test images, this corresponds to 332 blocks in the horizontal direction and 186 blocks in the vertical direction, each with 16 computational threads.
\begin{lstlisting}[
	caption=Grid configuration for the separable convolution device kernel,
	label=grid-configuration
]
#define TX 16
#define TY 16

@@

horizontal_convolve<<<grid_size_horizontal, block_size>>>(...);
vertical_convolve<<<grid_size_vertical, block_size>>>(...);
\end{lstlisting}
\par The convolution on the device executes only one serial loop across the dimensions of the filter exactly as it is defined in Equation \eqref{separable-convolution-formula}. This sum is then stored in the output array\footnote{In our implementation, we represent a two-dimensional image as a one-dimensional array by directly concatenating the rows of the matrix into a single array. Thus, for a matrix of size $M \times N$, the index $(i, j)$ would correspond to index $Ni + j$ in our one-dimensional array.} according to the block index, block size, and thread index. The implementation is modularized so that any separable filter can be applied to an input matrix.
\begin{lstlisting}[
	caption=Implementation of horizontal and vertical separated convolution,
	label=separable-convolution-implementation
]
__global__ void horizontal_convolve(int *d_out, int *x, int *h, int x_width, int x_height, int h_width, int h_height) {
    const int r = blockIdx.y * blockDim.y + threadIdx.y;
    const int c = blockIdx.x * blockDim.x + threadIdx.x;
    const int i = r * (x_width + h_width - 1) + c;

@@

    int sum = 0;
    for (int j = 0; j < h_height; j++) {
        int p = x_width*(r - j) + c;
        if (r - j >= 0 && r - j < x_height) {
            sum += h[j] * x[p];
        }
    }
    d_out[i] = (int)(constant_scalar * (double)sum);
    __syncthreads();
}
\end{lstlisting}
\par The first stage of our edge detection algorithm uses the separable convolution GPU implementation above to apply a Gaussian low-pass filter in Equation \ref{gaussian-filter} to remove high-frequency noise from the image that might be falsely labeled as an edge. Then, the same separable convolution kernel is used to apply the Sobel edge filter of Equations \ref{horizontal-sobel-filter} and \ref{vertical-sobel-filter} to the Gaussian low pass-filtered image in both the horizontal and vertical directions to approximate the horizontal and vertical partial derivatives, respectively. In order to reduce redundancy in computation, the final stage of non-maximum suppression and selective thresholding is separated into two tasks, each of which is implemented in parallel:
\begin{enumerate}
	\item Calculate the gradient magnitude matrix $\boldsymbol{G}$ and gradient angle matrix $\boldsymbol{\theta}$
	\item Implement Algorithm \ref{non-maximum-suppression-selective-thresholding-algorithm} with the inputs $\boldsymbol{G}$ and $\boldsymbol{\theta}$ calculated above
\end{enumerate}
\begin{lstlisting}[
	caption=GPU computation of gradient magnitude and angle,
	label=gradient-magnitude-angle-implementation
]
__global__ void gradient_magnitude_and_direction(
		double *dev_magnitude_output,
		double *dev_angle_output
		int input_width,
		int input_height,
		int *g_x,
		int *g_y
	) {
	const int r = blockIdx.y * blockDim.y + threadIdx.y;
	const int c = blockIdx.x * blockDim.x + threadIdx.x;
	const int i = r * (input_width + 2) + c;

	dev_magnitude_output[i] = sqrt(pow((double)g_x[i], 2) + pow((double)g_y[i], 2));
	dev_angle_output[i] = atan2((double)g_y[i], (double)g_x[i]);
}
\end{lstlisting}
Following completion of the computation of Listing \ref{gradient-magnitude-angle-implementation}, the result is input to our implementation of Algorithm \ref{non-maximum-suppression-selective-thresholding-algorithm}. While each of these two tasks operates in parallel (e.g. that each pixel's magnitude/angle or edge property is computed in parallel with the other pixels), the two tasks themselves execute serially. This is due to the fact that the non-maximum suppression and selective thresholding procedure relies on the gradient and angle at each pixel to have already been computed when the algorithm begins. Nonetheless, despite the fact that such serial separation of tasks reduces redundancies in gradient magnitude/angle computations\footnote{Computational redundancy arises from the fact that the non-maximum suppression and selective thresholding procedure needs access to the magnitude and angle of all the pixels in a box of area $\alpha^2$ centered at the current pixel. Calculating these properties directly for each pixel during the algorithm will result in calculation of the same magnitude and angle for the same pixel multiple times.}, further optimization is possible to reduce the amount of time that any thread is idle (which occurs after a particular pixel's gradient magnitude/angle is computed but before its edge property is computed).
\begin{lstlisting}[
	caption=GPU non-maximum suppression and selective thresholding,
	label=non-maximum-suppression-selective-thresholding-implementation
]
__global__ void thresholding_and_suppression(
	int *dev_output,
	double *dev_magnitude,
	double *dev_angle,
	int input_width,
	int input_height,
	int *g_x,
	int *g_y,
	int high_threshold,
	int low_threshold
) {
	const int r = blockIdx.y * blockDim.y + threadIdx.y;
	const int c = blockIdx.x * blockDim.x + threadIdx.x;
    const int i = r * (input_width + 2) + c;

@@

\section{Motion Detection Algorithm}
Our motion detection algorithm compares the edges from two frames temporally separated by a small interval of time and attempts to estimate regions of the frame that exhibit movement or motion. This process is then repeated continuously to provide a real-time visualization of the regions of movement in a continuous video stream. The motion detection algorithm we present is composed of the following steps:
\begin{enumerate}
	\item Given two frames $\boldsymbol{F}_1$ and $\boldsymbol{F}_2$ separated by a time interval of $\Delta t$, compute its edges $\boldsymbol{E}_1$ and $\boldsymbol{E}_2$.
	\item Create a strict binary difference matrix $\boldsymbol{D}_0$ between $\boldsymbol{E}_1$ and $\boldsymbol{E}_2$ whose value is high at locations at which an edge is present in frame but not the other, and low at locations at which both frames have an edge or no edge.
	\item Apply a parameter-controlled movement tolerance thresholding operation to the matrix $\boldsymbol{D}_0$ to obtain matrix $\boldsymbol{D}$, in order to suppress potential false positives of detected motion from incidental camera movement.
\end{enumerate}
\par $\boldsymbol{F}_1$ and $\boldsymbol{F}_2$ are obtained by sampling frames from a continuous video stream, a task aided by the OpenCV C library. Thus, the time interval $\Delta t$ will be small enough for a real-time algorithm. The edges $\boldsymbol{E}_1$ and $\boldsymbol{E}_2$ are computed with the edge detection algorithm of Section II. Below, we explore, in greater depth, the procedures for building a thresholded diffrence matrix and finally estimating the motion area by analyzing a spatial difference density map.

@@

\subsection{GPU Parallelization and CUDA Implementation}
Our CUDA implementation of our motion detection algorithm is similar to that for our implementation of separable convolution and edge detection. Using the same 16 threads per block, we divide the input image into the appropriate number of blocks such that the thresholded difference computation at each index $(i, j)$ of Algorithm \ref{difference-matrix-algorithm} has its own thread. Thus, each element of $\boldsymbol{D}$ is calculated in parallel.
\begin{lstlisting}[
	caption=GPU computation of thresholded difference matrix,
	label=difference-filter-implementation
]
__global__ void difference_filter(
	int *dev_out,
	int *edges_1,
	int *edges_2,
	int width,
	int height,
	int threshold
) {
	const int r = blockIdx.y * blockDim.y + threadIdx.y;
	const int c = blockIdx.x * blockDim.x + threadIdx.x;
	const int i = r * width + c;

    // Set it to 0 initially
    dev_out[i] = 0;
    int crop_size = 7;
    if (r > crop_size && c > crop_size && r < height - crop_size && c < width - crop_size && edges_1[i] != edges_2[i]) {
        // Set to 255 if there is a pixel mismatch
        dev_out[i] = 255;
        for (int x_apron = -threshold; x_apron <= threshold; x_apron++) {
            for (int y_apron = -threshold; y_apron <= threshold; y_apron++) {
                // Ensure the requested index is within bounds of image
                if (c + x_apron > 0 && r + y_apron > 0 && c + x_apron < width && r + y_apron < height) {
                    // Check if there is a matching pixel in the apron, within the threshold
                    if (edges_1[(r + y_apron) * width + c + x_apron] == edges_2[i]) {
                        // Set it back to 0 if a corresponding pixel exists within the vicinity of the match
                        dev_out[i] = 0;
                    }
                }
            }
        }
    }
}
\end{lstlisting}
The construction of the difference density matrix as in Algorithm \ref{difference-density-matrix-algorithm} is parallelized by considering the value of the thresholded difference matrix $\boldsymbol{D}'$ at each index $(r, c)$, mapping that index to the appropriate index in the difference density matrix $(i, j)$ where $i < m \wedge j < n$, then adding to the value of the array at $(i, j)$ at each encounter of a difference pixel in $\boldsymbol{D}'$. Our CUDA implementation therefore relies on the fact that the density array is initialized to 0 before the kernel is launched; this is accomplished in practice with a \texttt{calloc} operation on the host, followed by copying that array to GPU memory.
\begin{lstlisting}[
	caption=GPU computation of the difference density matrix,
	label=difference-density-implementation
]
__global__ void spatial_difference_density_map(
	double *density_map,
	int *difference,
	int width,
	int height,
	int horizontal_divisions,
	int vertical_divisions
) {
	int r = blockIdx.y * blockDim.y + threadIdx.y;
	int c = blockIdx.x * blockDim.x + threadIdx.x;
	int i = r * width + c;

@@

	const int scaling_factor = 1000;
	if (difference[i] != 0) {
		int i = (int)(vertical_divisions * r/(double)height);
		int j = (int)(horizontal_divisions * c/(double)width);
		density_map[i * horizontal_divisions + j] += scaling_factor/(double)block_size;
	}
}
\end{lstlisting}
The generation of the motion area estimation image is relatively straightforward: we map the index $(i, j)$ of $\boldsymbol{D}'$ back to an index $(r, c)$ where $r < M \wedge c < N$, and set the value at that index high or low accordingly.
\begin{lstlisting}[
	caption=GPU generation of the motion area estimation image,
	label=motion-area-estimation-implementation
]
__global__ void motion_area_estimate(
	int *motion_area,
	double *density_map,
	int width,
	int height,
	int horizontal_divisions,
	int vertical_divisions,
	double threshold
) {
	int r = blockIdx.y * blockDim.y + threadIdx.y;
	int c = blockIdx.x * blockDim.x + threadIdx.x;
	int i = r * width + c;

@@

\subsection{High Level Python Implementation}
Python is a widely used big-data programming language because of its extensive library support in such areas and its ease of use.

\section{GPU Speedup Results}
Our code for separable convolution, computation of the image's gradient magnitude and angle, non-maximum suppression and selective thresholding, and computation of the thresholded difference and spatial difference density matrices has both serial CPU and parallel CUDA implementations. Here, we present speedup results for each of these procedures on various image sizes obtained on a system with the following hardware:
\begin{itemize}
	\item Intel Core i5 4200U @ 2.6 GHz (2C, 4T)
	\item 8 GB of DDR3 memory @ 1600 MHz
	\item NVIDIA GeForce GTX 860M (1152 CUDA cores, 1029 MHz core clock, 2500 MHz memory clock, 2 GB GGDR5 video memory)
\end{itemize}

>>>>

\lstset{
	frame=tb,
	aboveskip=3mm,
	belowskip=3mm,
	showstringspaces=false,
	columns=flexible,
	basicstyle={\small\ttfamily},
	numbers=none,
	numberstyle=\tiny\color{gray},
	keywordstyle=\color{blue},
	commentstyle=\color{dkgreen},
	stringstyle=\color{mauve},
	breaklines=true,
	breakatwhitespace=true,
	tabsize=3
}

@@

\title{GPU Acceleration of Edge-Based Motion Detection and Machine Learning-Aided Facial Recognition with NVIDIA CUDA}

@@

\begin{abstract}
GPUs provide a powerful platform for parallel computations on large data inputs, namely images. In this paper, we explore a GPU-based implementation of a simplified adaptation of existing edge detection algorithms fast enough to operate on frames of a continuous video stream in real-time. We also demonstrate a practical application of edge detection--an edge-based method for motion detection estimation. Additionally, we explore the GPU-CPU speedup of existing OpenCV GPU computation libraries, namely, for facial recognition algorithms. Finally, we demonstrate the speedups as high as 10x we achieve with GPU parallelism, as compared to a reference serial CPU-based implementation.
\end{abstract}

\section{Introduction}
Graphics processing units (GPUs) are rapidly gaining popularity as a platform for parallelized computations on massive sets of data. Since much of the computations in image processing and computer vision are easily parallelized, graphics operations on GPUs achieve significant speedups compared to those done on their serial, CPU counterparts. Further, SDKs like the NVIDIA CUDA framework provide developers easy APIs to take advantage of the parallel computing power of GPUs. We take full advantage of the computational benefits of GPUs by implementing edge detection and motion detection algorithms in CUDA C, and making use of existing CUDA libraries for our facial recognition algorithm.
\par In this paper, we first detail the theory for our edge detection, motion detection, and facial recognition algorithms in Sections \ref{edge-detection-section}, \ref{motion-detection-section}, and \ref{facial-recognition-section}, respectively. At the end of Sections \ref{edge-detection-section} and \ref{motion-detection-section}, we describe our GPU code implementation of these algorithms with NVIDIA CUDA. At the end of Section \ref{facial-recognition-section}, we comment on the performance we achieve with a prebuilt, CUDA-based OpenCV GPU computation library, as opposed to that we achieve with a custom CUDA implementation as in Sections \ref{edge-detection-section} and \ref{motion-detection-section}. We present speedup results achieved with our CUDA implementation with respect to a reference serial, CPU implmentation in Section \ref{results}. Finally, we conclude in Section \ref{conclusion}.

\subsection{GPU Computation and the NVIDIA CUDA Framework}
Formally, CUDA (Compute Unified Device Architecture) is a parallel computing platform and programming model that exposes familiar C-based APIs for parallelized computations. CUDA is NVIDIA's platform for general-purpose computing on graphics processing units (commonly, GPGPU or GP$^2$U), or the use of a GPU for computations traditionally handled by the CPU. Generally, GPGPU is used to exploit the improved multithreaded performance and raw floating-point computational ability of GPUs over CPUs. For example, on modern hardware, an NVIDIA GeForce GTX 970 (1664 CUDA cores) exhibits peak single-precision floating point performance of nearly 3500 GFLOPS (floating-point operations per second), while an Intel Core i7 4790K (4C, 8T) achieves 100 GFLOPS. Our goal is to demonstrate the performance of GPU computing by solving a handful of existing problems in computer vision with CUDA: namely, edge detection, motion detection, and facial recognition.

\subsection{OpenCV and Python}
While CUDA is an excellent language for maximizing performance, it could be infeasible due to inexperience with C, time constraints, or lack of an NVIDIA GPU. To address these constraints, we also implemented our algorithms in Python using the Python OpenCV library. We chose Python because it is very easy for inexperienced programmers to use, and we chose OpenCV because it offers implementations of many common image processing algorithms. Unfortunately, using OpenCV limits us to the library's implementation, which does not currently support the use of GPU parallelization in Python without using C wrappers. For each of the algorithms in this paper, we present code snippets that provide a sample CPU-based Python implementation alongside a GPU-based parallel CUDA implementation.

\subsection{Motivation}
GPU parallelization is a powerful tool for accelerating image processing tasks. The motivation for our CUDA motion detection and facial recognition algorithms stems from their diverse applications and the ease of coding a GPU implementation, which is due in large part to the simplicity of CUDA APIs and OpenCV libraries. For example, motion detection is applied in real-time traffic analysis, vision-based security, and consumer technology (e.g. motion-triggered home automation solutions). Further, since our computer vision algorithm calculates the location of motion in a frame based on detected edges, our approach is both more accurate and supplemented with more specific data. Facial recognition is applied in real-time security, smart photographic analysis, and military purposes, including target-based missile guiding systems. Therefore, we view an opportunity use CUDA to explore the speedup possible by using a GPU to accelerate these common tasks in computer vision.

@@

\section{Edge Detection Algorithm}
\label{edge-detection-section}
Edge detection is the process of determining the locations of boundaries that separate objects, or sections of objects, in an image. This edge data simplifies the features of an image to only its basic outlines, which makes it a convenient input for many other problems in computer vision, including motion detection and tracking, object classification, three-dimensional reconstruction, and others. The edge detection algorithm we present is composed of three stages:
\begin{enumerate}
	\item Reduce the amount of high-frequency noise from the image using a two-dimensional low pass filter.
	\item Determine the two-dimensional gradient of the filtered image by applying partial derivatives in both the horizontal and vertical directions.
	\item Classify edges by suppressing surrounding pixels in the gradient direction whose magnitude is lesser, and applying selective thresholding to a limited apron around each selected pixel.
\end{enumerate}
Edges correspond to the points on the image that exhibit a high gradient magnitude. The edge data from our algorithm is then used as input to our motion detection algorithm.

@@

\subsection{GPU Parallelization and CUDA Implementation}
We begin with a parallelized implementation of separable convolution, since the same convolution operation is applied for both the initial Gaussian low pass filter and the edge detecting Sobel filter. Our parallel implementation involves first independently computing the convolution with the horizontal filter ($h_2$ in Equation \eqref{separable-convolution-definition}), then convolving that result with the vertical filter $h_1$. In computing the output $y[m, n]$ at every index $(i, j)$, we launch a kernel with 16 threads per block, where the total number of blocks along one dimension is equal to the size of that dimension divided by 16. For our 5312 x 2988 test images, this corresponds to 332 blocks in the horizontal direction and 186 blocks in the vertical direction, each with 16 computational threads.
\begin{lstlisting}[
	caption=Grid configuration for the separable convolution device kernel,
	label=grid-configuration,
	language=C
]
#define TX 16
#define TY 16

@@

horizontal_convolve<<<grid_size_horizontal, block_size>>>(...);
vertical_convolve<<<grid_size_vertical, block_size>>>(...);
\end{lstlisting}
\par The convolution on the device executes only one serial loop across the dimensions of the filter exactly as it is defined in Equation \eqref{separable-convolution-formula}. This sum is then stored in the output array\footnote{In our implementation, we represent a two-dimensional image as a one-dimensional array by directly concatenating the rows of the matrix into a single array. Thus, for a matrix of size $M \times N$, the index $(i, j)$ would correspond to index $Ni + j$ in our one-dimensional array.} according to the block index, block size, and thread index. The implementation is modularized so that any separable filter can be applied to an input matrix.
\begin{lstlisting}[
	caption=Implementation of horizontal and vertical separated convolution,
	label=separable-convolution-implementation,
	language=C
]
__global__ void horizontal_convolve(int *d_out, int *x, int *h, int x_width, int x_height, int h_width, int h_height) {
    const int r = blockIdx.y * blockDim.y + threadIdx.y;
    const int c = blockIdx.x * blockDim.x + threadIdx.x;
    const int i = r * (x_width + h_width - 1) + c;

@@

    int sum = 0;
    for (int j = 0; j < h_height; j++) {
        int p = x_width*(r - j) + c;
        if (r - j >= 0 && r - j < x_height) {
            sum += h[j] * x[p];
        }
    }
    d_out[i] = (int)(constant_scalar * (double)sum);
    __syncthreads();
}
\end{lstlisting}
\par The first stage of our edge detection algorithm uses the separable convolution GPU implementation above to apply a Gaussian low-pass filter in Equation \ref{gaussian-filter} to remove high-frequency noise from the image that might be falsely labeled as an edge. Then, the same separable convolution kernel is used to apply the Sobel edge filter of Equations \ref{horizontal-sobel-filter} and \ref{vertical-sobel-filter} to the Gaussian low pass-filtered image in both the horizontal and vertical directions to approximate the horizontal and vertical partial derivatives, respectively. In order to reduce redundancy in computation, the final stage of non-maximum suppression and selective thresholding is separated into two tasks, each of which is implemented in parallel:
\begin{enumerate}
	\item Calculate the gradient magnitude matrix $\boldsymbol{G}$ and gradient angle matrix $\boldsymbol{\theta}$
	\item Implement Algorithm \ref{non-maximum-suppression-selective-thresholding-algorithm} with the inputs $\boldsymbol{G}$ and $\boldsymbol{\theta}$ calculated above
\end{enumerate}
\begin{lstlisting}[
	caption=GPU computation of gradient magnitude and angle,
	label=gradient-magnitude-angle-implementation,
	language=C
]
__global__ void gradient_magnitude_and_direction(
		double *dev_magnitude_output,
		double *dev_angle_output
		int input_width,
		int input_height,
		int *g_x,
		int *g_y
	) {
	const int r = blockIdx.y * blockDim.y + threadIdx.y;
	const int c = blockIdx.x * blockDim.x + threadIdx.x;
	const int i = r * (input_width + 2) + c;

	dev_magnitude_output[i] = sqrt(pow((double)g_x[i], 2) + pow((double)g_y[i], 2));
	dev_angle_output[i] = atan2((double)g_y[i], (double)g_x[i]);
}
\end{lstlisting}
Following completion of the computation of Listing \ref{gradient-magnitude-angle-implementation}, the result is input to our implementation of Algorithm \ref{non-maximum-suppression-selective-thresholding-algorithm}. While each of these two tasks operates in parallel (e.g. that each pixel's magnitude/angle or edge property is computed in parallel with the other pixels), the two tasks themselves execute serially. This is due to the fact that the non-maximum suppression and selective thresholding procedure relies on the gradient and angle at each pixel to have already been computed when the algorithm begins. Nonetheless, despite the fact that such serial separation of tasks reduces redundancies in gradient magnitude/angle computations\footnote{Computational redundancy arises from the fact that the non-maximum suppression and selective thresholding procedure needs access to the magnitude and angle of all the pixels in a box of area $\alpha^2$ centered at the current pixel. Calculating these properties directly for each pixel during the algorithm will result in calculation of the same magnitude and angle for the same pixel multiple times.}, further optimization is possible to reduce the amount of time that any thread is idle (which occurs after a particular pixel's gradient magnitude/angle is computed but before its edge property is computed).
\begin{lstlisting}[
	caption=GPU non-maximum suppression and selective thresholding,
	label=non-maximum-suppression-selective-thresholding-implementation,
	language=C
]
__global__ void thresholding_and_suppression(
	int *dev_output,
	double *dev_magnitude,
	double *dev_angle,
	int input_width,
	int input_height,
	int *g_x,
	int *g_y,
	int high_threshold,
	int low_threshold
) {
	const int r = blockIdx.y * blockDim.y + threadIdx.y;
	const int c = blockIdx.x * blockDim.x + threadIdx.x;
    const int i = r * (input_width + 2) + c;

@@

\subsection{High-Level Python Implementation}
The Python OpenCV library exposes an API for the entire Canny edge detection algorithm. The Canny algorithm is similar to, but more nuanced than, our algorithm, substituting our selective thresholding procedure for hysteresis. However, the end results are very similar. As is evident from Listing \ref{python-edge-detection-implementation}, the procedure is very straightforward.
\begin{lstlisting}[
	caption=Python implementation of Canny edge detection with OpenCV,
	label=python-edge-detection-implementation,
	language=Python
]
LOW_THRESHOLD = 100
HIGH_THRESHOLD = 200
edges = cv2.Canny(img, LOW_THRESHOLD, HIGH_THRESHOLD)
\end{lstlisting}

\section{Motion Detection Algorithm}
\label{motion-detection-section}
Our motion detection algorithm compares the edges from two frames temporally separated by a small interval of time and attempts to estimate regions of the frame that exhibit movement or motion. This process is then repeated continuously to provide a real-time visualization of the regions of movement in a continuous video stream. The motion detection algorithm we present is composed of the following steps:
\begin{enumerate}
	\item Given two frames $\boldsymbol{F}_1$ and $\boldsymbol{F}_2$ separated by a time interval of $\Delta t$, compute its edges $\boldsymbol{E}_1$ and $\boldsymbol{E}_2$.
	\item Create a strict binary difference matrix $\boldsymbol{D}_0$ between $\boldsymbol{E}_1$ and $\boldsymbol{E}_2$ whose value is high at locations at which an edge is present in frame but not the other, and low at locations at which both frames have an edge or no edge.
	\item Apply a parameter-controlled movement tolerance thresholding operation to the matrix $\boldsymbol{D}_0$ to obtain matrix $\boldsymbol{D}$, in order to suppress potential false positives of detected motion from incidental camera movement.
\end{enumerate}
\par $\boldsymbol{F}_1$ and $\boldsymbol{F}_2$ are obtained by sampling frames from a continuous video stream, a task aided by the OpenCV C library. Thus, the time interval $\Delta t$ will be small enough for a real-time algorithm. The edges $\boldsymbol{E}_1$ and $\boldsymbol{E}_2$ are computed with the edge detection algorithm of Section II. Below, we explore, in greater depth, the procedures for building a thresholded diffrence matrix and finally estimating the motion area by analyzing a spatial difference density map.

@@

\subsection{GPU Parallelization and CUDA Implementation}
Our CUDA implementation of our motion detection algorithm is similar to that for our implementation of separable convolution and edge detection. Using the same 16 threads per block, we divide the input image into the appropriate number of blocks such that the thresholded difference computation at each index $(i, j)$ of Algorithm \ref{difference-matrix-algorithm} has its own thread. Thus, each element of $\boldsymbol{D}$ is calculated in parallel.
\begin{lstlisting}[
	caption=GPU computation of thresholded difference matrix,
	label=difference-filter-implementation,
	language=C
]
__global__ void difference_filter(
	int *dev_out,
	int *edges_1,
	int *edges_2,
	int width,
	int height,
	int threshold
) {
	const int r = blockIdx.y * blockDim.y + threadIdx.y;
	const int c = blockIdx.x * blockDim.x + threadIdx.x;
	const int i = r * width + c;

    // Set it to 0 initially
    dev_out[i] = 0;
    int crop_size = 7;
    if (r > crop_size && c > crop_size && r < height - crop_size && c < width - crop_size && edges_1[i] != edges_2[i]) {
        // Set to 255 if there is a pixel mismatch
        dev_out[i] = 255;
        for (int x_apron = -threshold; x_apron <= threshold; x_apron++) {
            for (int y_apron = -threshold; y_apron <= threshold; y_apron++) {
                // Ensure the requested index is within bounds of image
                if (c + x_apron > 0 && r + y_apron > 0 && c + x_apron < width && r + y_apron < height) {
                    // Check if there is a matching pixel in the apron, within the threshold
                    if (edges_1[(r + y_apron) * width + c + x_apron] == edges_2[i]) {
                        // Set it back to 0 if a corresponding pixel exists within the vicinity of the match
                        dev_out[i] = 0;
                    }
                }
            }
        }
    }
}
\end{lstlisting}
The construction of the difference density matrix as in Algorithm \ref{difference-density-matrix-algorithm} is parallelized by considering the value of the thresholded difference matrix $\boldsymbol{D}'$ at each index $(r, c)$, mapping that index to the appropriate index in the difference density matrix $(i, j)$ where $i < m \wedge j < n$, then adding to the value of the array at $(i, j)$ at each encounter of a difference pixel in $\boldsymbol{D}'$. Our CUDA implementation therefore relies on the fact that the density array is initialized to 0 before the kernel is launched; this is accomplished in practice with a \texttt{calloc} operation on the host, followed by copying that array to GPU memory.
\begin{lstlisting}[
	caption=GPU computation of the difference density matrix,
	label=difference-density-implementation,
	language=C
]
__global__ void spatial_difference_density_map(
	double *density_map,
	int *difference,
	int width,
	int height,
	int horizontal_divisions,
	int vertical_divisions
) {
	int r = blockIdx.y * blockDim.y + threadIdx.y;
	int c = blockIdx.x * blockDim.x + threadIdx.x;
	int i = r * width + c;

@@

	const int scaling_factor = 1000;
	if (difference[i] != 0) {
		int i = (int)(vertical_divisions * r/(double)height);
		int j = (int)(horizontal_divisions * c/(double)width);
		density_map[i * horizontal_divisions + j] += scaling_factor/(double)block_size;
	}
}
\end{lstlisting}
The generation of the motion area estimation image is relatively straightforward: we map the index $(i, j)$ of $\boldsymbol{D}'$ back to an index $(r, c)$ where $r < M \wedge c < N$, and set the value at that index high or low accordingly.
\begin{lstlisting}[
	caption=GPU generation of the motion area estimation image,
	label=motion-area-estimation-implementation,
	language=C
]
__global__ void motion_area_estimate(
	int *motion_area,
	double *density_map,
	int width,
	int height,
	int horizontal_divisions,
	int vertical_divisions,
	double threshold
) {
	int r = blockIdx.y * blockDim.y + threadIdx.y;
	int c = blockIdx.x * blockDim.x + threadIdx.x;
	int i = r * width + c;

@@

\subsection{High-Level Python Implementation}
The Python implementation of Algorithm \ref{difference-density-matrix-algorithm} is very straightforward and looks almost exactly like the pseudo-code, making it easy to implement even for those with minimal programming experience.
\begin{lstlisting}[
	caption=Python implementation of the motion area estimation image,
	label=python-motion-area-estimattion-implementation,
	language=Python
]
import numpy as np

def detect_motion(e1, e2):
    height, width, depth = np.shape(e1)
    d = difference(e1, e2)
    d_prime = np.zeros((M, N), np.uint8)
    for i in range(M):
        for j in range(N):
            x = 0
            for i_prime in range(height/M):
                for j_prime in range(width/N):
                    if d[i*height/M+i_prime, j*width/N+j_prime] != 0:
                        x += 1
            if M*N*x/(height*width*1.0) > T:
                d_prime[i, j] = 255
    return d_prime
\end{lstlisting}

\section{Facial Recognition Algorithm}
\label{facial-recognition-section}
Our facial recognition implementation is based off of the HAAR features-based cascade classifiers method proposed by Viola-Jones in 2001. It's split into two phases: training and detection. The algorithm is as follows:
\begin{enumerate}
	\item Extract HAAR features of a face from a set of positive and negative training images. 
	\item Use adaptive boosting machine learning to train a multi-stage, or cascade, classifier against positive and negative HAAR features. 
	\item Apply a final cascade classifier to a loaded image or real time video frame. 
\end{enumerate}
Due to the resource capacity necessary to train a full facial recognition classifier, we used a pre-trained frontal face cascade classifier and used OpenCV's object detection library, which encapsulates the algorithm above, and whose theory detailed in the following sections.
